BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications

### 2016-12-27 Paper2/BOAST.tex
Ported Fred modifications to IEEE style.

\newcommand{\Specfem}{\productname{Specfem 3D}\xspace}
\newcommand{\Cuda}{\productname{Cuda}\xspace}
\newcommand{\OCL}{\productname{OpenCL}\xspace}
\newcommand{\code}[1]{\texttt{#1}}

@@

\markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for IEEE Journals}
%\journal{Journal of Parallel and Distributed Computing}

%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%%

%% Numbered
%\bibliographystyle{model1-num-names}

%% Numbered without titles
%\bibliographystyle{model1a-num-names}

%% Harvard
%\bibliographystyle{model2-names.bst}\biboptions{authoryear}

%% Vancouver numbered
%\usepackage{numcompress}\bibliographystyle{model3-num-names}

%% Vancouver name/year
%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}

%% APA style
%\bibliographystyle{model5-names}\biboptions{authoryear}

%% AMA style
%\usepackage{numcompress}\bibliographystyle{model6-num-names}

%% `Elsevier LaTeX' style
\bibliographystyle{IEEEtran}
%%%%%%%%%%%%%%%%%%%%%%%

@@

%\begin{frontmatter}

@@

%% or include affiliations in footnotes:
\author{Brice Videau, Kevin Pouget, Luigi Genovese, Thierry Deutsch, Dimitri Komatitsch, Frédéric Desprez, Jean-François Méhaut}

%\address[mymainaddress]{LIG/CNRS}

%\begin{keyword}
%Code Generation \sep Portability \sep Genericity \sep Productivity and Software
%Design \sep High Performance Computing \sep Autotuning \sep Non Regression
%Testing
%\end{keyword}

@@

\begin{abstract}
The Abstract.
\end{abstract}

\section{Introduction}

Porting and tuning HPC applications to new platforms is tedious and costly in term of human resources.
Many developer teams are more concentrated on solving the physical problems faster and better with new algorithms
and simulating new physics. 
This is why optimization even if it is important is a task which is sometimes neglected or hidden for many developers.
In the same way, the domain specific language (DSL) has exactly this goal,  the complete separation of the high-level code dedicated to simulate new physics
and the low-level one which is more dependent to the HPC architecture. 
Nevertheless, decoupling the high and low levels don't avoid the need of optimization which becomes more and more unavoidable 
with the new architectures based on many cores and accelerators.

Unfortunately, portability efforts of the low level i.e. computing kernels are often lost when migrating to a new
architecture. Worse, code may lose maintainability because several versions of
some functionalities coexist, usually with a lot of duplication.
Thus productivity of porting and tuning efforts is low as a huge fraction of
those developments are never used after the platform they were intended for is
decommissioned.

The primary goal of the metaprogramming framework BOAST is to simplify the optimization of the computing kernels for HPC
applications. The main idea is to use a high level language to define an API of computing kernels i.e. subroutines and
their related arguments, but also describe the way to optimize it (unrolling, vectorization, parametrization, templating, tiling).
BOAST will test automatically the many possibilities of optimizations and choose the best one after validation from a
reference version.
Then BOAST can generate pieces of Fortran, C, OpenCL or CUDA codes, test if for a specific target architecture.

With the definition of an API, BOAST can hide the details of the implementation at high level but also is the way to
factorize different kernels depending only on parameters as datatypes, the length of filters or the boundary conditions.
The maintainability and genericity of HPC codes is often limited. One of the reason is that producing
generic code in Fortran 90/95 is difficult as the language does not really fit
for it. Sometimes, adding genericity or functionalities can be quite costly and degrades performance as optimization
opportunities that come from over-specification are lost.
BOAST can provide a solution to handle genericity and add new functionalities at low levels.
For instance, the optimization of short convolutions depends strongly
on the length of the filters. This is why HPC codes using finite differences handle only one order. BOAST allows
handling any order and even more generalizing to another numerical scheme.

In this paper, we illustrate the optimization problems related to scientific computing with the Laplace equation.
Then we present our solution, BOAST, its architecture and its functionalities. We apply BOAST to three use cases, the
Laplace equation, short multi-dimensional convolutions to perform differential operators and finally
the wavelet transform which are used in a quantum chemistry code, BigDFT.
Then BOAST is applied to a full application, SPECFEM3D dedicated to the seismic wave propagation simulator.
80 GPU computing kernels have been ported improving considerably the maintanability and providing a single code base for
OpenCL and CUDA kernel versions.
Finally we compare BOAST with other related works.

% Porting and tuning HPC applications to new platforms is tedious and costly in
% terms of human resources. Nonetheless, it is a very important aspect of the
% Mont-Blanc project. Indeed, for the project, more than ten applications were
% selected to be ported and optimized for the prototype platform.
% 
% Unfortunately, portability efforts are often lost when migrating to a new
% architecture. Worse, code may lose maintainability because several versions of
% some functionalities coexist, usually with a lot of duplication.
% 
% Thus productivity of porting and tuning efforts is low as a huge fraction of
% those developments are never used after the platform they were intended for is
% decommissioned.
% 
% Genericity of HPC codes is often limited. One of the reason is that producing
% generic code in Fortran 90/95 is difficult as the language does not really fit
% for it. Sometimes, adding genericity degrades performance as optimization
% opportunities that come from over-specification are lost.
% 
% Functionality of HPC codes is tied to the previous point. Without genericity,
% adding new functionalities can be quite costly.

\section{Background and Motivation}

The motivation of the BOAST project comes from the need of the HPC communities which have full and large applications
difficult to port and maintain for emerging architectures based on accelerators, longer vector lengths, or low power
processors.

  \subsection{Evolution of HPC Architectures}

Evolution of HPC Architectures is rapid and also diverse: in the last 5
years no less than 6 architectures have been number one in the Top500:
\begin{itemize}
\item Intel Processor + Xeon Phi (Tianhe-2)
\item AMD Processor + NVIDIA GPU (Titan)
\item IBM BlueGene/Q (Sequoia)
\item Fujitsu SPARC64 (K computer)
\item Intel Processor + NVIDIA GPU (Tianhe-1)
\item AMD Processor (Jaguar)
\end{itemize}
Being able to efficiently use those architectures on such a small
time-frame is challenging.

The race to exascale is not going to simplify the environment. All of the above
architectures can be considered. Network architectures also can be very
diverse.  For instance European FP7 project DEEP considers using Accelerators
(XEON Phi) while the European FP7 project Mont-Blanc considers using low-power
embedded processor with integrated GPU.

Running existing applications on those new architectures is an open research
subject as well as an ongoing porting effort.  Thus, those projects have work
packages dedicated to applications. Those work packages are dedicated to porting
and optimizing Scientific applications on those new architectures. In the DEEP
project six applications were selected for porting and optimizing, eleven were
selected in the Mont-Blanc project.

  \subsection{Scientific Computing Applications}

Scientific Computing Applications are usually developed by physicist, chemist or
meteorologist. Those codes are usually written in Fortran for historical and
performance reasons. Codes can be quite huge (several thousands lines of code [LOC])
with lots of functionalities. Nonetheless, they are usually based on computing
kernels.  Computing kernels are resource intensive and well-defined parts of a
program that usually work on precisely defined data. Those kernels represent the most
time-consuming part of an HPC application, and consequently, they are the prime target for
optimization.

Those applications are often developed by several individuals. Sometimes, some
of those developers only work a few months on the application. Maintaining
optimized code written by someone else is quite a challenge. Several languages
and programming paradigms can also be used in a project and thus the maintainer
must be knowledgeable in several areas of expertise. Portability problems can
also be caused by the availability of an optimizing compiler for a specific
language and architecture. C compilers are usually the first available while
Fortran may sometimes arrive later.

@@

  \subsection{How Should Computing Kernels be Written?}

The problematic here is to obtain computing kernels that present good
performances on the architectures encountered by the application while still
being portable after the optimization process took place. Indeed the
application might encounter one of the many architectures that can be found in
HPC. Investing manpower to optimize the application for a new architecture is
reasonable, suffering hindrance from previous optimization work is not. Thus
optimizations have to be as orthogonal as possible from one another so as to be
easily activated and deactivated.

@@

What we want is computing kernels that are written:
\begin{itemize}
\item in a \emph{portable} manner,
\item in a way that raises developer \emph{productivity},
\item and, present good \emph{performance}.
\end{itemize}

\section{Basic Example: OpenCL Laplace}
\label{sec:laplace}

During one of the Mont-Blanc face to face meeting a talk on OpenCL optimization
on the Mali GPU was given. One of the case study was a Laplace filter on the
GPU~\cite{opencl_arm_training}.

This is a good example of an algorithm that is simple in its formulation but
can be complex to optimize because many different optimizations are available
and can be combined together. The correct combination of optimizations will
depend on the target architecture.

\subsection{The Laplace Filter}

Listing~\ref{lst:laplace_c_filter} shows a simple implementation of the Laplace
filter in C99. For clarity, boundary conditions have been omitted.

\lstset{style=customc,
        caption={Laplace C99 Filter},
        breaklines=true,
        breakautoindent=false,
        label={lst:laplace_c_filter}}
\begin{lstlisting}
void laplace(const int width, const int height,
             const unsigned char src[height][width][3],
                   unsigned char dst[height][width][3]){
  for (int j = 1; j < height-1; j++) {
    for (int i = 1; i < width-1; i++) {
      for (int c = 0; c < 3; c++) {
        int tmp =
   -src[j-1][i-1][c] -   src[j-1][i][c] - src[j-1][i+1][c]\
  - src[j  ][i-1][c] + 9*src[j  ][i][c] - src[j  ][i+1][c]\
  - src[j+1][i-1][c] -   src[j+1][i][c] - src[j+1][i+1][c];
        dst[j][i][c] = (tmp<0 ? 0 : (tmp>255 ? 255 : tmp));
      }
    }
  }
}
\end{lstlisting}

@@

\lstset{style=customcl,
        caption={Laplace OpenCL Filter},
        breaklines=true,
        breakautoindent=false,
        label={lst:laplace_opencl_filter}}
\begin{lstlisting}
kernel laplace(const int width, const int height,
               global const uchar *src,
               global       uchar *dst){
  int i = get_global_id(0);
  int j = get_global_id(1);
  for (int c = 0; c < 3; c++) {
    int tmp = -src[3*width*(j-1) + 3*(i-1) + c]\
             - src[3*width*(j-1) + 3*(i  ) + c]\
             - src[3*width*(j-1) + 3*(i+1) + c]\
             - src[3*width*(j  ) + 3*(i-1) + c]\
           + 9*src[3*width*(j  ) + 3*(i  ) + c]\
             - src[3*width*(j  ) + 3*(i+1) + c]\
             - src[3*width*(j+1) + 3*(i-1) + c]\
             - src[3*width*(j+1) + 3*(i  ) + c]\
             - src[3*width*(j+1) + 3*(i+1) + c];
    dst[3*width*j + 3*i + c] = clamp(tmp, 0, 255);
  }
}
\end{lstlisting}

\subsection{Possible Optimizations of the OpenCL Version}

@@

\lstset{style=customcl,
        caption={Laplace OpenCL Filter Vectorized with Synthesized Loads},
        breaklines=true,
        breakautoindent=false,
        label={lst:laplace_opencl_filter_vector_synth}}
\begin{lstlisting}
  uchar16 v11_ =vload16(0, src + 3*width*(j-1) + 3*5*i -3);
  uchar16 v13_ =vload16(0, src + 3*width*(j-1) + 3*5*i +3);
  uchar16 v12_ =uchar16(v11_.s3456789a, v13_.s56789abc);
  uchar16 v21_ =vload16(0, src + 3*width*(j  ) + 3*5*i -3);
  uchar16 v23_ =vload16(0, src + 3*width*(j  ) + 3*5*i +3);
  uchar16 v22_ =uchar16(v21_.s3456789a, v23_.s56789abc);
  uchar16 v31_ =vload16(0, src + 3*width*(j+1) + 3*5*i -3);
  uchar16 v33_ =vload16(0, src + 3*width*(j+1) + 3*5*i +3);
  uchar16 v32_ =uchar16(v31_.s3456789a, v33_.s56789abc);
\end{lstlisting}

\paragraph{Temporary Variables Size} Using \textit{int} to store intermediary
results is unnecessary. Listing~\ref{lst:laplace_opencl_filter_short} show how
the code is modified to use smaller types. This yields a speedup of 1.3 for most
image sizes.

\lstset{style=customcl,
        caption={Laplace OpenCL Filter Vectorized Using \textit{short}},
        breaklines=true,
        breakautoindent=false,
        label={lst:laplace_opencl_filter_short}}
\begin{lstlisting}
  short16 v11 = convert_short16(v11_);
  short16 v12 = convert_short16(v12_);
  short16 v13 = convert_short16(v13_);
  short16 v21 = convert_short16(v21_);
  short16 v22 = convert_short16(v22_);
  short16 v23 = convert_short16(v23_);
  short16 v31 = convert_short16(v31_);
  short16 v32 = convert_short16(v32_);
  short16 v33 = convert_short16(v33_);

@@

Nonetheless, the process is frustrating. The optimizations are never evaluated
independently from one another. Some were arbitrarily configured (the number of
pixels chosen for instance). Testing the different combinations of those
optimizations and the different parameters would be very costly in developer
time and, moreover, like all repetitive and tedious tasks, error prone. Thus
every created kernel would have to be thoroughly tested to ensure no error was
done.

In the next Section we will present our answer to these issues: BOAST.  BOAST is
a framework dedicated to kernel description, optimization, regression testing
and autotuning.

\section{BOAST: Using Code Generation in Application Autotuning}

@@

%  \begin{itemize}
%  \item Abstraction of programming concepts: variables, procedures (
%portability, productivity )
%  \item Operator overloading: simpler syntax ( productivity, portability )
%  \item Two languages in one ( productivity, performance ) 
%  \end{itemize}

@@

%\begin{figure}
%\begin{center}
%\includegraphics[scale=0.8]{BOAST_DSL.pdf}
%\caption{BOAST DSL intent.}
%\label{fig:dsl}
%\end{center}
%\end{figure}

@@

\begin{figure}
\lstset{style=BOAST,
        caption={BOAST code}, 
        label={lst:BOAST_algebra},
        numbers=right }
\begin{lstlisting}
i = BOAST::Int("i") # or BOAST::Variable::new("i", Int)
k = BOAST::Int("k", :size => 8)
l = BOAST::Real("l", :dim => [ BOAST::Dim(-5, 21) ], :local => true )
BOAST::decl i, k, l
BOAST::pr i === 5
j = i + 5
BOAST::pr k === j * 2
BOAST::pr l[k] === 1.0
BOAST::register_funccall("sin")
BOAST::pr l[k+1] === BOAST::sin(j)
\end{lstlisting}

@@

\begin{figure}
\lstset{style=BOAST,
        caption={BOAST code}, 
        label={lst:BOAST_procedure},
        numbers=right }
\begin{lstlisting}
n = Int("n", :dir=>:in)
ndat = Int("ndat", :dir=>:in)
x = Real("x", :dir=>:in, :dim=>[Dim(0, n-1), Dim(ndat)])
y = Real("y", :dir=>:out, :dim=>[Dim(ndat), Dim(0, n-1)])
p = Procedure("magicfilter", [n, ndat, x, y])
opn p
close p
\end{lstlisting}

@@

%Considering source-to-source transformations that we investigate with the BOAST
%tool in this paper, we can compare ourselves to POET~\cite{POET2012}  and
%ABCLibScript~\cite{ABCLib2006}. However, both tools apply transformations
%to the syntaxic tree of a program while BOAST intervenes well before in the
%transformation process. Indeed, one may first use BOAST on order to generate a
%program (a specific syntactic tree) and then benefit from the transformation
%facilities offered by one of the above tools.

@@

In this section we will present the different aspects of BOAST run-time that
allow this automation. Those aspects include:
  \begin{itemize}
  \item multi target language generation ( performance, portability )
  \item compilation ( productivity, performance )
  \item execution ( productivity, performance )
  \item tracer, dumper and replay for non regression tests ( productivity )
  \end{itemize}

  \subsubsection{Multi-target Language Generation}
\label{sec:multitarget}
Language availability and performance varies between platforms. It is thus
important to express computing kernels in different languages, based on the
availability and their respective merits on the target platform. Some languages
have additional features, such as languages that target GPUs (OpenCL, CUDA), or
languages that support threading paradigms (OpenMP). The developer should be
given tools to determine what kind of language is currently used in order to be
able to use those additional features.

@@

Nvidia \Cuda and \OCL are based on the same programming model: a
massively parallel accelerator running in disjoint, non addressable
memory environment. Thanks to that proximity, we have been able to carry
out most of the porting task with only a limited knowledge about
\Specfem internal physics.

This lack of \Specfem internal knowledge led us to be particularly
careful to the path we undertook for the port, as we would have
been unable to understand how and why the application was not
operating properly, if it was to fail.

Hence, our first milestone in the port was the translation of
\Specfem's \Cuda kernels into BOAST EDSL. This way, we could ask BOAST
framework to generate a CUDA version of the kernels, plug them back
into \Specfem and get (after fixing compile-time errors---prototypes
and naming mistakes mainly) a first set of \Specfem seismograms.

As we had expected, the seismograms were erroneous. But with the help
of shell scripts and BOAST framework ability to store and provide the
kernels' original source code, we built a set of \Specfem binaries
including only \emph{one} BOAST-generated kernel, with all others
reference-kernels. Running and validating all these binaries enabled
use to pinpoint the misbehaving kernels. We finished the debugging
with a side-by-side comparison that highlighted the coding mistakes.

\paragraph{Porting run-time to \OCL} The second part of the port
consisted in the translation of the CPU-side of the application, from
\Cuda API to \OCL API. Most of the functions of the interfaces are
very similar, with only naming-convention and data-structure
distinctions. Hence, it was clear that automatic rewriting tools
(namely \code{sed} regexp and \code{emacs-lisp} functions) could be
useful. To give an idea of the cost of a \emph{manual} rewriting, we
can count (in \# of \OCL API function calls): 70 kernel ``function
calls'', 790 arguments to set, 230 memory transfers, 160 buffer
creations, and 270 releases.

>>>>

\newcommand{\code}[1]{\texttt{#1}}
\newcommand*{\Num}{N\textsuperscript{o}\xspace}

@@

\bibliographystyle{IEEEtran}

@@

@@

%% or include affiliations in footnotes:
\author{
   \IEEEauthorblockN{Brice Videau\IEEEauthorrefmark{1}\IEEEauthorrefmark{2}, Kevin Pouget\IEEEauthorrefmark{1}\IEEEauthorrefmark{3}, Luigi Genovese\IEEEauthorrefmark{4}, Thierry Deutsch\IEEEauthorrefmark{4}, Dimitri Komatitsch\IEEEauthorrefmark{2}, Frédéric Desprez\IEEEauthorrefmark{1}\IEEEauthorrefmark{5}, Jean-François Méhaut\IEEEauthorrefmark{1}\IEEEauthorrefmark{3}}\\
   \IEEEauthorblockA{\IEEEauthorrefmark{1}Laboratoire d'Informatique de Grenoble}\\
   \IEEEauthorblockA{\IEEEauthorrefmark{2}Centre National de la Recherche Scientifique}\\
   \IEEEauthorblockA{\IEEEauthorrefmark{3}Université Joseph Fourier}\\
   \IEEEauthorblockA{\IEEEauthorrefmark{4}CEA/INAC/SP2M/L\_Sim}\\
   \IEEEauthorblockA{\IEEEauthorrefmark{5}Inria}
}

\markboth{International Journal of Scientific Computing}
{Videau \MakeLowercase{\textit{et al.}}: BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications}
%\address[mymainaddress]{LIG/CNRS}

%\begin{keyword}
%  Code Generation \sep Portability \sep Genericity \sep Autotuning \sep Non Regression Testing \sep Productivity and Software Design \sep High Performance Computing
%\end{keyword}

@@

\begin{abstract}
The portability of real HPC applications on new platforms is an open and very
delicate problem.  Especially the performance portability of the underlying
computing kernels is problematic as they need to be tuned for each and every
platform the application encounters. This paper presents BOAST, a
metaprogramming framework dedicated to computing kernels.  BOAST allows the
description of a kernel and its possible optimizations using a DSL. BOAST
runtime will then compare the different versions' performance as well as verify
their exactness. BOAST is applied to three use cases: a Laplace kernel in
OpenCL and two HPC applications BigDFT (electronic density computation) and
SPECFEM3D (seismic and wave propagation).
\end{abstract}

%%% INTRODUCTION %%%
\section{Introduction}

Porting and tuning HPC applications to new platforms is tedious and costly in
term of human resources.  Many developer teams are more concentrated on solving
the physical problems faster and better with new algorithms and simulating new
physics. This is why optimization, while important, is a task which is
sometimes neglected or hidden for many developers.  In the same way, Domain
Specific Language (DSLs) have exactly this goal, the complete separation of the
high-level code dedicated to simulate new physics and the low-level one which
is more dependent on the HPC architecture. Nevertheless, decoupling the high
and low levels does not avoid the need of optimizations which become more and
more unavoidable with new architectures based on many cores and accelerators.

Unfortunately, portability efforts of the low level codes (i.e. computing
kernels) are often lost when migrating to a new architecture. Worse, code may
lose maintainability because several versions of some functionalities coexist,
usually with a lot of duplication.  Thus productivity of porting and tuning
efforts is low as a huge fraction of those developments are never used after
the platform they were intended for is decommissioned.

The primary goal of the BOAST metaprogramming framework is to simplify the
optimization of computing kernels for HPC applications. The main idea is to use
a high level language to define an API of computing kernels i.e. subroutines
and their related arguments, but also describe the way to optimize it
(unrolling, vectorization, parametrization, templating, tiling). BOAST will
then test automatically the many possibilities of optimizations and choose the
best one after validation from a reference version. Then BOAST can generate
pieces of Fortran, C, OpenCL, or CUDA codes, test them for a specific target
architecture.

With the definition of an API, BOAST can hide the implementation details at
high level but also is a way to factorize different kernels depending only on
parameters such as data types, length of filters, or boundary conditions. The
maintainability and genericity of HPC codes is often limited. One of the reason
is that producing generic code in Fortran 90/95 is difficult as the language
does not really fit for it. Sometimes, adding genericity or functionalities can
be quite costly and may degrade performance as optimization opportunities that
come from over-specification are lost. BOAST can provide a solution to handle
genericity and add new functionalities at low levels.

The paper is organized as follows. First in Section~\ref{sec:backg} we further
motivate the need for metaprogramming frameworks.  Secondly in
Section~\ref{sec:laplace} we illustrate the optimization problems related to
scientific computing with the Laplace equation. Thirdly we present our
solution, called BOAST, its architecture and its functionalities in
Section~\ref{sec:boast}. Fourthly we apply BOAST to three use cases: the
Laplace equation motivating example, short multi-dimensional convolutions from
the BigDFT quantum chemistry application, and the seismic wave propagation
simulator SPECFEM3D where 80 GPU computing kernels have been ported improving
considerably the maintainability and providing a single code base for OpenCL
and CUDA kernel versions. Finally, before a conclusion and future works, we
compare BOAST with other related works in Section~\ref{sec:related}.

%%% BACKGROUND AND MOTIVATION %%%
\section{Background and Motivation\label{sec:backg}}

The motivation of the BOAST project comes from the need of the HPC communities
which have full and large applications.  These applications are difficult to
port and maintain for emerging architectures especially since new architectures
are based on accelerators, longer vector units, or low power processors.

%%% EVOLUTION OF HPC ARCHITECTURES %%%
\subsection{Evolution of HPC Architectures}

Evolution of HPC Architectures is rapid and also diverse. In the last seven
years no less than 7 architectures have been number one in the
Top500~\cite{webtop500}:
\begin{itemize}
\item Sunway processor (TaihuLight, 2016)
\item Intel Processor + Xeon Phi (Tianhe-2, 2013)
\item AMD Processor + NVIDIA GPU (Titan, 2012)
\item IBM BlueGene/Q (Sequoia, 2012)
\item Fujitsu SPARC64 (K computer, 2011)
\item Intel Processor + NVIDIA GPU (Tianhe-1, 2010)
\item AMD Processor (Jaguar, 2009)
\end{itemize}

Being able to efficiently use those architectures on such a small period is
challenging for scientists and application developers.

The race to exascale is not going to simplify the environment. All of the above
machine architectures can be considered to build these supercomputers and other
will appear soon. For instance European FP7 project DEEP considers using
Accelerators (XEON Phi) while the European FP7 project Mont-Blanc considers
using low-power embedded processor with integrated GPU.

Running existing applications on these new architectures is an open research
subject as well as an ongoing porting effort.  Thus, those projects have work
packages dedicated to applications. Those work packages are dedicated to
porting and optimizing Scientific applications on those new architectures. In
the DEEP project six applications were selected for porting and optimizing,
eleven were selected in the Mont-Blanc project.

%%% SCIENTIFIC COMPUTING APPLICATIONS %%%
\subsection{Scientific Computing Applications}

Scientific Computing Applications are usually developed by other science
researchers such as physicists, chemists or meteorologists. Those codes are
usually written in Fortran for historical and performance reasons. Codes can be
quite huge (several thousands Lines of Code, LoC) with lots of functionalities.
Nonetheless, they are usually based on computing kernels.  Computing kernels
are resource intensive and well-defined parts of a program that usually work on
precisely defined data. Those kernels represent the most time-consuming part of
an HPC application, and consequently, they are the prime target for
optimizations.

Those applications are often developed by several individuals. Sometimes, some
of those developers only work a few months on the application. Maintaining
optimized code written by someone else is quite a challenge. Several languages
and programming paradigms can also be used in a project and thus the maintainer
must be knowledgeable in several areas of expertise. Portability problems can
also be caused by the availability of an optimizing compiler for a specific
language and architecture. C compilers are usually first available while
Fortran may sometimes come later.

@@

%%% HOW SHOULD COMPUTING KERNELS BE WRITTEN? %%%
\subsection{How Should Computing Kernels be Written?}

The problematic here is to obtain computing kernels that present good
performance on the architectures encountered by the application while staying
portable after the optimization process took place. Indeed the application
might encounter one of the many architectures that can be found in HPC.
Investing manpower to optimize the application for a new architecture is
reasonable, suffering hindrance from previous optimization work is not. Thus
optimizations have to be as orthogonal as possible from one another so as to be
easily activated and deactivated.

@@

What we want is computing kernels that are written in a \emph{portable} manner,
in a way that raises developer \emph{productivity}, and they must present good
\emph{performance}.

%%% BASIC EXAMPLE: OPENCL LAPLACE %%%
\section{Basic Example: OpenCL Laplace\label{sec:laplace}}

During one of the Mont-Blanc face to face meeting a talk on OpenCL optimization
on the Mali GPU was given. One of the case study was a Laplace
filter~\cite{opencl_arm_training}. This is a good example of an algorithm that
is simple in its formulation but can be complex to optimize because many
different optimizations are available and can be combined together. The correct
combination of optimizations will depend on the target architecture.

%%% THE LAPLACE FILTER %%%
\subsection{The Laplace Filter}

Listing~\ref{lst:laplace_c_filter} shows a simple implementation of the Laplace
filter in C99. For sake of clarity, boundary conditions have been omitted.

\lstset{style=customc,
        caption={Laplace C99 Filter.},
        breaklines=true,
        breakautoindent=false,
        label={lst:laplace_c_filter}}
\begin{lstlisting}
void laplace(const int width, const int height,
             const unsigned char src[height][width][3],
                   unsigned char dst[height][width][3]){
  for (int j = 1; j < height-1; j++) {
    for (int i = 1; i < width-1; i++) {
      for (int c = 0; c < 3; c++) {
        int tmp =
   -src[j-1][i-1][c] -   src[j-1][i][c] - src[j-1][i+1][c]\
  - src[j  ][i-1][c] + 9*src[j  ][i][c] - src[j  ][i+1][c]\
  - src[j+1][i-1][c] -   src[j+1][i][c] - src[j+1][i+1][c];
        dst[j][i][c] = (tmp<0 ? 0 : (tmp>255 ? 255 : tmp));
      }
    }
  }
}
\end{lstlisting}

@@

\lstset{style=customcl,
        caption={Laplace OpenCL Filter.},
        breaklines=true,
        breakautoindent=false,
        label={lst:laplace_opencl_filter}}
\begin{lstlisting}
kernel laplace(const int width, const int height,
               global const uchar *src,
               global       uchar *dst){
  int i = get_global_id(0);
  int j = get_global_id(1);
  for (int c = 0; c < 3; c++) {
    int tmp = -src[3*width*(j-1) + 3*(i-1) + c]\
             - src[3*width*(j-1) + 3*(i  ) + c]\
             - src[3*width*(j-1) + 3*(i+1) + c]\
             - src[3*width*(j  ) + 3*(i-1) + c]\
           + 9*src[3*width*(j  ) + 3*(i  ) + c]\
             - src[3*width*(j  ) + 3*(i+1) + c]\
             - src[3*width*(j+1) + 3*(i-1) + c]\
             - src[3*width*(j+1) + 3*(i  ) + c]\
             - src[3*width*(j+1) + 3*(i+1) + c];
    dst[3*width*j + 3*i + c] = clamp(tmp, 0, 255);
  }
}
\end{lstlisting}

%%% POSSIBLE OPTIMIZATIONS OF THE OPENCL VERSION %%%
\subsection{Possible Optimizations of the OpenCL Version}

@@

\lstset{style=customcl,
        caption={Laplace OpenCL Filter Vectorized with Synthesized Loads.},
        breaklines=true,
        breakautoindent=false,
        label={lst:laplace_opencl_filter_vector_synth}}
\begin{lstlisting}
  uchar16 v11_ =vload16(0, src + 3*width*(j-1) + 3*5*i -3);
  uchar16 v13_ =vload16(0, src + 3*width*(j-1) + 3*5*i +3);
  uchar16 v12_ =uchar16(v11_.s3456789a, v13_.s56789abc);
  uchar16 v21_ =vload16(0, src + 3*width*(j  ) + 3*5*i -3);
  uchar16 v23_ =vload16(0, src + 3*width*(j  ) + 3*5*i +3);
  uchar16 v22_ =uchar16(v21_.s3456789a, v23_.s56789abc);
  uchar16 v31_ =vload16(0, src + 3*width*(j+1) + 3*5*i -3);
  uchar16 v33_ =vload16(0, src + 3*width*(j+1) + 3*5*i +3);
  uchar16 v32_ =uchar16(v31_.s3456789a, v33_.s56789abc);
\end{lstlisting}

\paragraph{Temporary Variables Size} Using \emph{int} to store intermediary
results is unnecessary. Listing~\ref{lst:laplace_opencl_filter_short} show how
the code is modified to use smaller types. This yields a speedup of 1.3 for
most image sizes.

\lstset{style=customcl,
        caption={Laplace OpenCL Filter Vectorized Using \emph{short}.},
        breaklines=true,
        breakautoindent=false,
        label={lst:laplace_opencl_filter_short}}
\begin{lstlisting}
  short16 v11 = convert_short16(v11_);
  short16 v12 = convert_short16(v12_);
  short16 v13 = convert_short16(v13_);
  short16 v21 = convert_short16(v21_);
  short16 v22 = convert_short16(v22_);
  short16 v23 = convert_short16(v23_);
  short16 v31 = convert_short16(v31_);
  short16 v32 = convert_short16(v32_);
  short16 v33 = convert_short16(v33_);

@@

Nonetheless, the process is frustrating. The optimizations are never evaluated
independently from one another. Some were arbitrarily configured (like the
number of pixels chosen for instance). Testing the different combinations of
those optimizations and the different parameters would be very costly in
developer time and, moreover, like all repetitive and tedious tasks, error
prone. Thus every created kernel would have to be thoroughly tested to ensure
no error was done.

In the next Section we will present our answer to these issues: BOAST.  BOAST
is a framework dedicated to kernel description, optimization, regression
testing, and autotuning.

%%% BOAST: USING CODE GENERATION IN APPLICATION AUTOTUNING %%%
\section{BOAST: Using Code Generation in Application Autotuning\label{sec:boast}}

@@

@@

@@

\begin{figure}
\lstset{style=BOAST,
        caption={BOAST code.}, 
        label={lst:BOAST_algebra},
        numbers=right }
\begin{lstlisting}
i = BOAST::Int("i") # or BOAST::Variable::new("i", Int)
k = BOAST::Int("k", :size => 8)
l = BOAST::Real("l", :dim => [ BOAST::Dim(-5, 21) ], :local => true )
BOAST::decl i, k, l
BOAST::pr i === 5
j = i + 5
BOAST::pr k === j * 2
BOAST::pr l[k] === 1.0
BOAST::register_funccall("sin")
BOAST::pr l[k+1] === BOAST::sin(j)
\end{lstlisting}

@@

\begin{figure}
\lstset{style=BOAST,
        caption={BOAST code.}, 
        label={lst:BOAST_procedure},
        numbers=right }
\begin{lstlisting}
n = Int("n", :dir=>:in)
ndat = Int("ndat", :dir=>:in)
x = Real("x", :dir=>:in, :dim=>[Dim(0, n-1), Dim(ndat)])
y = Real("y", :dir=>:out, :dim=>[Dim(ndat), Dim(0, n-1)])
p = Procedure("magicfilter", [n, ndat, x, y])
opn p
close p
\end{lstlisting}

@@

@@

In this section we will present the different aspects of BOAST run-time that
allow this automation. Those aspects are multi-target language generation
(\emph{performance}, \emph{portability}), kernel compilation
(\emph{productivity}, \emph{performance}), kernel execution
(\emph{productivity}, \emph{performance}) and last, kernel tracer, dumper and
replay for non regression tests (\emph{productivity}).

  \subsubsection{Multi-target Language Generation\label{sec:multitarget}}
Language availability and performance varies between platforms. It is thus
important to express computing kernels in different languages, based on the
availability and their respective merits on the target platform. Some languages
have additional features, such as languages that target GPUs (OpenCL, CUDA), or
languages that support threading paradigms (OpenMP). The developer should be
given tools to determine what kind of language is currently used in order to be
able to use those additional features.

@@

Nvidia CUDA and OpenCL are based on the same programming model: a
massively parallel accelerator running in disjoint, non addressable
memory environment. Thanks to that proximity, we have been able to carry
out most of the porting task with only a limited knowledge about
SPECFEM3D internal physics.

This lack of SPECFEM3D internal knowledge led us to be particularly
careful to the path we undertook for the port, as we would have
been unable to understand how and why the application was not
operating properly, if it was to fail.

Hence, our first milestone in the port was the translation of
SPECFEM3D's CUDA kernels into BOAST EDSL. This way, we could ask BOAST
framework to generate a CUDA version of the kernels, plug them back
into SPECFEM3D and get (after fixing compile-time errors---prototypes
and naming mistakes mainly) a first set of SPECFEM3D seismograms.

As we had expected, the seismograms were erroneous. But with the help
of shell scripts and BOAST framework ability to store and provide the
kernels' original source code, we built a set of SPECFEM3D binaries
including only \emph{one} BOAST-generated kernel, with all others
reference-kernels. Running and validating all these binaries enabled
use to pinpoint the misbehaving kernels. We finished the debugging
with a side-by-side comparison that highlighted the coding mistakes.

\paragraph{Porting run-time to OpenCL} The second part of the port
consisted in the translation of the CPU-side of the application, from
CUDA API to OpenCL API. Most of the functions of the interfaces are
very similar, with only naming-convention and data-structure
distinctions. Hence, it was clear that automatic rewriting tools
(namely \code{sed} regexp and \code{emacs-lisp} functions) could be
useful. To give an idea of the cost of a \emph{manual} rewriting, we
can count (in \# of OpenCL API function calls): 70 kernel ``function
calls'', 790 arguments to set, 230 memory transfers, 160 buffer
creations, and 270 releases.
